A Methodology and Implementation of Automated Emissions Harmonization for Use in Integrated Assessment Models

### 2018-01-31 paper.tex
addressed all text reviewer comments

\glspl{iam} are tools used to understand the complex interactions between
energy, economy, land use, water, and climate systems. \glspl{iam} provide
global projections of systemic change by dividing the world into a number of
representative regions (typically 10 to 30), the definition of which is distinct
for each model \cite{krey_global_2014}. Results from \glspl{iam} are integral in
a number of international studies, which notably include projections of climate
and economic futures. Recently, the \gls{iam} community has developed scenarios
based on the \glspl{ssp} \cite{van_vuuren_energy_2017, fricko_marker_2017,
  fujimori_ssp3:_2017, calvin_ssp4:_2017, kriegler_fossil-fueled_2017} which
quantify a variety of potential global futures. The \glspl{ssp} are designed to
be used in research that include \gls{esm} simulations, climate impact,
adaptation and climate mitigation studies \cite{vuuren_new_2013}.

While \glspl{iam} are implemented in myriad ways, including simulation and
optimization, the core inputs and outputs are similar across different
models. Modeling teams incorporate data on energy systems, land use, economics,
demographics and emissions sources and concentrations, among other data, in
order to provide consistent existing trajectories of modeled variables. The
models then provide estimates of future trajectories of these variables under
various socio-economic and technological assumptions as well as proposed policy
constraints, e.g., targets for future \gls{ghg} emissions.

@@

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{decision_tree.pdf}
    \caption[]{
      \label{fig:decision_tree}
      The default method decision tree used in the \code{aneris} software
      library. For all decisions, upper (purple) branches represent a ``yes''
      response and lower (orange) branches represent a ``no'' response. The
      coefficient of variation, $c_v$, is defined in Equation \ref{eqs:cov},
      $dH$ is defined as $\left|\frac{h(t_i) - m(t_i)}{h(t_i)}\right|$, and
      decision-making thresholds for $c_v$ and $dH$ are described below.  Where
      present, convergence years of default methods are provided below the
      method name in parentheses.  Methods labeled in \textit{green} are likely
      to closely match unharmonized results, methods in \textit{yellow} will
      likely somewhat match unharmonized results, and methods in \textit{red}
      can be expected to have a large relative difference between harmonized and
      unharmonized results.}
  \end{center}
\end{figure}

A number of characteristics impact the decision of which default method to
select based on the effect of the characteristic on the potential harmonized
trajectory. For example, it is possible for models to report zero values in the
harmonization year in situations in which technologies are introduced in future
time periods in regions or for sectors which produce an emissions species that
is absent in the initial modeling period. In such cases, an offset method is
required as a ratio method would mask future emissions and erroneously harmonize
the trajectory.

In most cases, however, models do report values in the harmonization
year. Figure \ref{fig:cases} displays a number of illustrative example
trajectories which highlight the possible issues resulting from harmonizing
model results in different contexts. Panels \code{a} and \code{b} highlight
examples where model results peak mid-century, behavior that is seen in a number
of scenarios with general emissions mitigation effects, such as pollution
controls applied by developing nations on transport and industry sectors. Panel
\code{a} highlights a case where model base-year values and history are relative
close whereas Panel \code{b} shows a situation where model values and history
are relatively far apart. Panels \code{c} and \code{d} show similar model
trajectories that peak mid-century but also have negative emissions. Models can
report negative emissions in contexts in which a gas species which can be
extracted from the environment and stored, as is the case for \cotwo in future
scenarios with climate mitigation enacted via the deployment of carbon capture
and sequestration technologies or direct carbon capture technologies. Again, the
relative difference with historical values differ between the panels to explore
harmonization method choices in each situation.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{cases.pdf}
    \caption[]{
      \label{fig:cases}
      The effect of different harmonization routines on model trajectories under
      ``normal'' circumstances (Panel \code{a}), when there is a large
      difference between historical and model values in the harmonization year
      (Panels \code{b} and \code{d}), and when model trajectories result in
      negative emissions by the end of the modeling time horizon (Panels
      \code{c} and \code{d}). Identical model trajectories are used in each row
      (Panels \code{a}, \code{b}; \code{c}, \code{d}). In each column,
      historical values are increased in the base year by an order of magnitude
      (from 10 to 100). In each Panel, a subset of the potential routines
      provide a better harmonization quality than others as described in the
      text.
    }
  \end{center}
\end{figure}

@@

The value of $c_v$ is then tested against a threshold, $\tau_{c_v}$. To
determine this threshold, an analysis of the recent \gls{ceds} and \gls{luc} historical data
has been performed. Figure \ref{fig:cov} shows the distribution of \gls{luc} $c_v$s
and non-\gls{luc} $c_v$s as determined for historical data aggregated to the model
regions of 5 different \glspl{iam} involved in the \gls{ssp} process: AIM-CGE, IMAGE, GCAM4,
MESSAGE-GLOBIOM, and REMIND-MAGPIE \TODO{add refs}, each of which have varying
definitions of native model regions comprising different collections of
countries. Therefore, each data point comprising Figure \ref{fig:cov} represents
a realization of $c_v$ for a single combination of native model region, sector,
and emissions species \footnote{A full listing of all sectors and species is
  presented in the case-study discussion in Section \ref{sec:results}, Table
  \ref{tab:sp}}. A threshold value of $\tau_{c_v} = 20$ has been chosen based on
these observations as it optimally divides the two distributions. Importantly,
tails of the \gls{luc} and non-\gls{luc} overlap, thus there are both false positives (~7\%
of non-\gls{luc} trajectories) and false negatives (~10\% of \gls{luc}
trajectories). However, as any regional definition is model dependent and thus
any regional aggregation is possible an automated detection mechanism is
necessary.

>>>>

\glspl{iam} are tools used to understand the complex interactions between
energy, economy, land use, water, and climate systems. \glspl{iam} provide
global projections of systemic change by dividing the world into a number of
representative regions (typically 10 to 30), the definition of which is distinct
for each model \cite{krey_global_2014}. Results from \glspl{iam} are integral in
a number of international studies, which notably include projections of climate
and economic futures. Recently, the \gls{iam} community has developed scenarios
based on the \glspl{ssp} \cite{fujimori_ssp3:_2017, van_vuuren_energy_2017,
  calvin_ssp4:_2017, fricko_marker_2017, kriegler_fossil-fueled_2017} which
quantify a variety of potential global futures. The \glspl{ssp} are designed to
be used in research that include \gls{esm} simulations, climate impact,
adaptation and climate mitigation studies \cite{vuuren_new_2013}.

While \glspl{iam} are implemented in myriad ways\footnote{IAM models are
  numerous and have a long history in the scientific literature. Various IAMs
  have collaborated to produce
  \href{http://themasites.pbl.nl/models/advance/index.php/ADVANCE_wiki}{community
    IAM documentation} which readers can access for a full treatment of model
  implementation and features.}, including simulation and optimization, the core
inputs and outputs are similar across different models. Modeling teams
incorporate data on energy systems, land use, economics, demographics and
emissions sources and concentrations, among other data, in order to provide
consistent existing trajectories of modeled variables. The models then provide
estimates of future trajectories of these variables under various socio-economic
and technological assumptions as well as proposed policy constraints, e.g.,
targets for future \gls{ghg} emissions.

@@

\subsection{The Conceptual Basis for Choosing Harmonization Methods}\label{sec:concept}

The goals of any harmonization exercise is threefold: aligning model results in
the harmonization year to a common historical data source, faithfully
representing the original \gls{iam}’s internal consistency between the driver
of emissions (e.g. energy use) and emissions, and maintaining critical
parameters from the original scenario design. All harmonization methods achieve
the first goal by design. If the difference between the model base year and
historical values are small, considering the second and third goals leads to a
\code{reduce\_ratio} method that converges prior to the final model year as the
natural best choice. It preserves the ratio of \gls{iam} output and emissions
inventory in the base year while also matching the original model output at some
point in the modeled time period. It furthermore maintains the consistency of
the model's usage of energy technology, volume of agricultural activities, and
abatement options with harmonized emissions trajectory.

However, other concerns may lead to a better-informed choice than using a
blanket \code{reduce\_ratio} method for all emissions trajectories. For example,
emissions from \gls{luluc} are known to have high year-to-year variation, and
therefore historical data may change drastically depending on the base year
considered. In such a situation, a \code{reduce\_ratio} that converges at a year
past the modeled time period is a better choice in order to smooth out
discrepancies between the historical data used to develop model and the new data
source being used for harmonization.

Separately, if there are large discrepancies between the model results in the
base year and the historic data used for harmonization, convergence methods can
result in harmonized trajectories that do not faithfully represent the
underlying drivers of emissions. Furthermore, if models report negative
emissions, as is possible in scenarios designed to depict the deployment of
climate mitigation policies with large \cotwo reductions and eventual storage,
then end-of-century emissions characteristics can be of chief import in order to
faithfully match the design parameters of the original scenarios, such as global
mean temperature and other climate metrics.

Accordingly, we have developed a \textit{decision tree} approach to
harmonization method choice, discussed in Section \ref{sec:tree}, in order to
balance each of these concerns and use cases in a robust and automated manner.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{decision_tree.pdf}
    \caption[]{
      \label{fig:decision_tree}
      The default method decision tree used in the \code{aneris} software
      library. For all decisions, upper (purple) branches represent a ``yes''
      response and lower (orange) branches represent a ``no'' response. The
      coefficient of variation, $c_v$, is defined in Equation \ref{eqs:cov},
      $dH$ is defined as $\left|\frac{h(t_i) - m(t_i)}{h(t_i)}\right|$, and
      decision-making thresholds for $c_v$ and $dH$ are described below.  Where
      present, convergence years of default methods are provided below the
      method name in parentheses.  Convergence years are chosen in order to
      balance the three harmonization goals discussed in Section
      \ref{sec:concept}. Methods labeled in \textit{green} are likely to closely
      match unharmonized results, methods in \textit{yellow} will likely
      somewhat match unharmonized results, and methods in \textit{red} can be
      expected to have a large relative difference between harmonized and
      unharmonized results.}
  \end{center}
\end{figure}

A number of characteristics impact the decision of which default method to
select based on the effect of the characteristic on the potential harmonized
trajectory. For example, it is rare but possible for models to report zero
values in the harmonization year in situations in which technologies are
introduced in future time periods in regions or for sectors which produce an
emissions species that is absent in the initial modeling period. In such cases,
an offset method is required as a ratio method would mask future emissions and
erroneously harmonize the trajectory.

In most cases, however, models do report values in the harmonization
year. Figure \ref{fig:cases} displays a number of example trajectories which
highlight the possible issues resulting from harmonizing model results in
different contexts. These trajectories do not correspond to specific model
results; rather they serve as illustrative examples of the kinds of trajectories
observed in practice. Panels \code{a} and \code{b} highlight examples where
model results peak mid-century, behavior that is seen in a number of scenarios
with general emissions mitigation effects, such as pollution controls applied by
developing nations on transport and industry sectors. Panel \code{a} highlights
a case where model base-year values and history are relative close whereas Panel
\code{b} shows a situation where model values and history are relatively far
apart. Panels \code{c} and \code{d} show similar model trajectories that peak
mid-century but also have negative emissions. Models can report negative
emissions in contexts in which a gas species which can be extracted from the
environment and stored, as is the case for \cotwo in future scenarios with
climate mitigation enacted via the deployment of carbon capture and
sequestration technologies or direct carbon capture technologies. Again, the
relative difference with historical values differ between the panels to explore
harmonization method choices in each situation.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\textwidth]{cases.pdf}
    \caption[]{
      \label{fig:cases}
      We present here a number of illustrative examples of the effect of
      different harmonization routines on model trajectories under ``normal''
      circumstances (Panel \code{a}), when there is a large difference between
      historical and model values in the harmonization year (Panels \code{b} and
      \code{d}), and when model trajectories result in negative emissions by the
      end of the modeling time horizon (Panels \code{c} and \code{d}). Identical
      model trajectories are used in each row (Panels \code{a}, \code{b};
      \code{c}, \code{d}). In each column, historical values are increased in
      the base year by an order of magnitude (from 10 to 100). In each Panel, a
      subset of the potential routines provide a better harmonization quality
      than others as described in the text.  }
  \end{center}
\end{figure}

@@

The value of $c_v$ is then tested against a threshold, $\tau_{c_v}$. To
determine this threshold, an analysis of the recent \gls{ceds} and \gls{luc}
historical data has been performed. Figure \ref{fig:cov} shows the distribution
of \gls{luc} $c_v$s and non-\gls{luc} $c_v$s as determined for historical data
aggregated to the model regions of 5 different \glspl{iam} involved in the
\gls{ssp} process: AIM-CGE \cite{fujimori_ssp3:_2017},
IMAGE\cite{van_vuuren_energy_2017}, GCAM4\cite{calvin_ssp4:_2017},
MESSAGE-GLOBIOM\cite{fricko_marker_2017}, and
REMIND-MAGPIE\cite{kriegler_fossil-fueled_2017}, each of which have varying
definitions of native model regions comprising different collections of
countries. Therefore, each data point comprising Figure \ref{fig:cov} represents
a realization of $c_v$ for a single combination of native model region, sector,
and emissions species \footnote{A full listing of all sectors and species is
  presented in the case-study discussion in Section \ref{sec:results}, Table
  \ref{tab:sp}}. A threshold value of $\tau_{c_v} = 20$ has been chosen based on
these observations as it optimally divides the two distributions. Importantly,
tails of the \gls{luc} and non-\gls{luc} overlap, thus there are both false
positives (~7\% of non-\gls{luc} trajectories) and false negatives (~10\% of
\gls{luc} trajectories). However, as any regional definition is model dependent
and thus any regional aggregation is possible an automated detection mechanism
is necessary.
